Schema Evolution in Interactive Programming Systems

### 2024-08-23 paper.tex
Type->schema

The first challenge problem that we present in terms of our conceptual framework is live programming
of user interfaces based on the Elm architecture. In this model, a reactive web application is
structured in terms of current state and events that affect the state. The implementation then
consists of two functions called \texttt{update} and \texttt{render}:

@@

\noindent
The programming model works as follows:

@@

\noindent
Some systems that use the Elm architecture support hot reloading when the
implementation of the \texttt{render} or \texttt{update} function changes, but they typically
discard state and restart the application when the \texttt{State} type changes. A more effective
live programming system would allow live updates to the structure of the two types too.

@@

\frameworkbox{
\begin{wrapfigure}{r}{12em}
\vspace{0em}
\includegraphics[width=12em]{figures/arr-steps.png}
\vspace{0em}
\end{wrapfigure}
\frameworkboxtitle{Primitive Type Transformations}
In this case study, the evolving type is the \texttt{State} type. When the type evolves,
the program logic (code) and the current live value of the type (data) needs to evolve
correspondingly. The challenge is easier if we see the evolution as a sequence of more primitive
transformations, such as those in Table~\ref{tbl:elmchanges}, that each has a corresponding
primitive code and data transformation. The primitive transformations may either be invoked
through a user interface, or inferred from textual code edits made by the programmer.
}

@@

\section{Entity Evolution in Data-Centric Systems}
If we see programming systems as stateful environments where programs co-exist with their persistent
data, we can gain valuable insights from research on database systems.
Although schema evolution is well-studied in databases (\S\ref{sec:related} Related Work), past work
does not resolve all problems in our second challenge.

@@

\begin{figure}
\begin{adjustwidth}{-2em}{-3em}
\begin{subfigure}[b]{35em}\vspace{0pt}
  \sffamily
  \small
  \begin{tabular}{ |r|l|r|r|l|l|}
     \hline
     oid & item & quantity & ship\_date & customer\_name & customer\_address \\
     \hline \hline
     1 & Anvil & 1 & 2/3/23 & Wile E Coyote & 123 Desert Station \\
     \hline
     2 & Dynamite & 2 & & Daffy Duck & White Rock Lake \\
     \hline
     3 & Bird Seed & 1 & & Wile E Coyote & 123 Desert Station \\
     \hline
  \end{tabular}
  \caption{Spreadsheet recording orders with all associated information}
  \label{fig:db-orders}
\end{subfigure}
\hfill
\\[-0.5em]~\\
\begin{subfigure}[b]{20em}\vspace{0pt}
  \sffamily
  \small
  \begin{tabular}{ |r|l|l|}
    \hline
    cid & customer\_name & customer\_address \\
    \hline \hline
    1 & Wile E Coyote & 123 Desert Station \\
    \hline
    2 & Daffy Duck & White Rock Lake \\
    \hline
  \end{tabular}
  \caption{Customers table after type evolution}
  \label{fig:db-customers}
\end{subfigure}
\hfill
\begin{subfigure}[b]{20em}\vspace{0pt}
  \sffamily
  \small
  \begin{tabular}{ |r|l|r|r|r|}
     \hline
     oid & item & quantity & ship\_date & cid \\
     \hline \hline
     1 & Anvil & 1 & 2/3/23 & 1 \\
     \hline
     2 & Dynamite & 2 & & 2  \\
     \hline
     3 & Bird Seed & 1 & & 1 \\
     \hline
    \end{tabular}
  \caption{Orders linking to Customers, after type evolution}
 \label{fig:db-orders-link}
\end{subfigure}
\end{adjustwidth}
\vspace{0.25em}
\caption{Types and data before and after performing Extract Entity}
\label{fig:db}
\end{figure}

@@

\frameworkbox{
\begin{wrapfigure}{r}{12em}
\vspace{-0.5em}
\includegraphics[width=12em]{figures/arr-db.png}
\vspace{-1em}
\end{wrapfigure}
\frameworkboxtitle{Equivalence in Schema Evolution}
We described our conceptual framework in terms of types, code and data, but this case
study shows that the same patterns appear elsewhere. In database systems, the schema
$S_1$ changes into $S_2$ and the data $D_1$ need to transform to corresponding $D_2$.
Our framing shows that queries $Q_1$ also need to transform, which is not always addressed
in work on schema evolution. Extract/Absorb Entity operations are also interesting because
they transform an equivalent pair of schemas and so it should be possible to apply them
(and the corresponding data and query transformations) automatically.
}

@@

\section{Evolution of Computational Documents}
A different kind of type evolution is present in programming systems based on documents.
The idea, dating back to Boxer \cite{diSessa86}, is to represent programs as documents
with embedded computations. In our model, document can contain simple formulas (akin to those
in Potluck~\cite{Litt2023} or spreadsheets) that may refer to other nodes of the document.
We assume that documents can be concurrently edited by multiple users in a local-first manner,
i.e., local edits are later synchronized with peers.

@@

   \vspace{0.5em}
   \textbf{Conference budget}\\
   Travel cost per speaker:\\
   \hspace*{2em} \$1200 \\
   Number of speakers: \\
   \hspace*{2em} \texttt{=COUNT(/table[id='speakers']/tbody/tr)} \\
   Travel expenses:\\
   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
   }}\caption{Geoffrey adopts changes from all three co-organizers.}
   \label{fig:conf-finfin}
\end{subfigure}
\end{adjustwidth}
\caption{
Initial version of the document (not shown) contains a list of three speakers (Goldberg, Hamilton,
Jennings). The first challenge involves merging a data edit
(Fig.~\ref{fig:conf-add}) with a type and data edit (Fig.~\ref{fig:conf-tab}); the second involves
merging a type and data edit (Fig.~\ref{fig:conf-tab}) with edit adding code (Fig.~\ref{fig:conf-budget}).
Formulas in the final version (Fig.~\ref{fig:conf-finfin}) are updated to match the new document structure.}
\label{fig:conf}
\end{figure}

@@

\begin{enumerate}
\item Jonathan adds an additional speaker to the list and sorts the list of speakers
  alphabetically by their first name. This changes data, but not the structure (type) of
  the document (Figure~\ref{fig:conf-add}).
\item Tomas refactors the list into a table. He splits the single textual value
  into a name and an email (using a comma as the separator) and adds an additional column for
  tracking which of the organizers should contact the speaker (Figure~\ref{fig:conf-tab}).
  This changes the document structure (implicit type).
\end{enumerate}

@@

\frameworkbox{
\begin{wrapfigure}{r}{13em}
\vspace{-0.5em}
\includegraphics[width=13em]{figures/arr-datafork.png}
\vspace{-1em}
\end{wrapfigure}
\frameworkboxtitle{Merging Data and Code Edits}
In this example, the \emph{type} is implicit and represents the structure of the document.
It involves two independent edits. In one (top right), the data $D_1$ is
changed to $D_2'$, but the type $T_1$ remains unchanged. In the other (bottom left), the
type $T_1$ and data $D_1$ co-evolve into $T_2$ and $D_2$ and the data is further changed
to $D_3$. The challenge is to merge the two edits and obtain jointly modified data $D_3'$
of type $T_2$. In the next section, we will see the same pattern, but also involving
code edits. We assume the \emph{convergence} model in this section, but will discuss
the more challenging \emph{divergence} model later in the paper.}

@@

The type and data evolution happens through a sequence of edits. Table~\ref{tbl:docchanges}
lists edits appearing in our examples (but not a complete list). A system may
use a structure editor that provides high-level commands for triggering those edits.
Recognising user intent from plain text editing of source code may make the challenge harder.

@@

\noindent
The challenge is, again, to merge the edits done independently by the co-organizers.
The challenge extends the previous one by adding code. The change to the document structure
(\emph{type}) thus has to co-evolve with both the \emph{data} and the \emph{code}.
As shown in Figure~\ref{fig:conf-finfin}, the system needs to change the equation
\texttt{COUNT(/ul[id='speakers']/li)} to \texttt{COUNT(/table[id='speakers']/tbody/tr)}.
If the edit is recorded as a list of high-level edits shown in Figure~\ref{tbl:docchanges}
(wrapping element, changing their types, etc.), then each edit needs to perform corresponding
edit to selectors in formulas in the document.

\begin{table}[t]
\begin{adjustwidth}{-2em}{-2em}
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Change & Tag of an element & Apply to all matching elements and affected formulas\\
Wrap & Selected element(s) & Wrap elements and update selectors in affected formulas  \\
Split & Selected element(s) & Split matching elements using a rule (regular expression)\\
Add & New item to a list & Add data without affecting document structure (type)\\
Reorder & List items & Reorder data without affecting document structure (type)\\
\bottomrule
\end{tabular}
\vspace{0.3em}
\caption{Changes to type and data and how the programming system should respond}
\label{tbl:docchanges}
\end{adjustwidth}
\vspace{-1em}
\end{table}

\subsection*{Remarks: Liveness and Divergence}
The above challenge focuses on the local-first software scenario, but it can be extended to also
apply to live programming. If formulas in the document are evaluated on-the-fly, the system
needs to identify edits that invalidate some of the values (by changing the data that a formula
depends on, or by changing the equation). It should then automatically re-evaluate or invalidate
the computed results (requiring explicit re-evaluation). In this scenario, we distinguish between
\emph{permanent data} (information in the document) and transient \emph{data} (computed by code).
Whereas permanent data should be synchronized and co-evolve with the document structure (type),
transient data can be discarded.

@@

\section{Divergence Control in Spreadsheets}
Type evolution becomes more challenging when data, code, and types cannot all be updated
together in an atomic way. In these situations, different versions must coexist and interoperate
with one another. For example, edits done to data of a new type must be applied to
data of an older type, or code written against older type must run against data of a newer type.

@@

As a concrete example, we add the problem of divergence control to the
earlier Extract Entity challenge. Every month the orders department sends its spreadsheet to
the accounting department, which copies the data to its own version of the spreadsheet.
This version uses the same structure (type), but includes additional code for financial tracking.
When the orders department migrates its spreadsheet to a database and extracts out customers,
the accounting department does not want to conform. They want to continue using their old
spreadsheet with their additional code.

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-1.5em}
\includegraphics[width=14em]{figures/arr-backport.png}
\vspace{-2.5em}
\end{wrapfigure}
\frameworkboxtitle{Bidirectional Data Migration}
The scenario adds the need for transferring data changes from a new type to an older type.
In the diagram on the right, we only show data and type, ignoring the code added by the
accounting department. We start with data $D_1$ of type~$T_1$. The accounting department
evolves the type (via Extract Entity), obtaining data $D_2$ of type $T_2$ using an automatically
generated forward transformation. It further modifies the data, obtaining $D_3$. When the
accounting department receives data $D_3$, it needs to migrate it back from type $T_2$ to
$T_1$ using an automatically generated backward transformation. That is,
type evolution needs to be accompanied with both forward and backward projection for data
(and code) migration.}

@@

\frameworkbox{
\begin{wrapfigure}{r}{14.5em}
\vspace{-0.5em}
\includegraphics[width=14.5em]{figures/arr-divergence.png}
\vspace{-1.5em}
\end{wrapfigure}
\frameworkboxtitle{Maintaining Divergent Variants}
The extra complexity in the above scenario arises from the fact that both types and
code are forked and evolve independently. In one branch, original types $T_1$ evolve
into $T_2$ and original code $C_1$ into $C_3$. In another branch, original types $T_1$
evolve into $T'_2$ and original code $C_1$ into $C'_3$. The user then wants to merge
the independently done changes into a new version of types and code, written as
$T_2 \diamond T'_2$ and $C_3 \diamond C'_3$. The diverging transformations may make
conflicting changes that would need to be resolved with assistance from the user.}

@@

The challenge of maintaining multiple divergent types in programming systems can
be well illustrated using \emph{multiplicity change}, which is a common kind of schema change.
In this case a field changes from storing
a single value to storing multiple values. For example, we might start by storing a single
address for each contact in an address book, before realizing that we need to store multiple
addresses for a single contact. Alternatively, we might assign each todo in a list to one person before
realizing we want the ability to assign a todo to multiple people.

In a relational schema, solving this problem might require extracting a normalized table for
the linked entity; some of the challenges of this approach are covered in the Extract Entity
challenge. In this section we will instead consider a type structure with support for arrays;
in this context, we can have a multiplicity change by turning a scalar value into an array value.
%
A change of multiplicity in the type requires producing a transformation for both the corresponding
data and the corresponding code. The challenge becomes particularly difficult to handle when there
is ongoing divergence between multiple versions of the types and data.

\challenge{Multiplicity Change in Types}

@@

\frameworkbox{
\begin{wrapfigure}{r}{13em}
\vspace{-1em}
\includegraphics[width=13em]{figures/arr-backport.png}
\vspace{-2em}
\end{wrapfigure}
\frameworkboxtitle{Bidirectional Data Migration (Revisited)}
The diagram illustrating the scenario remains the same as in the case of divergence control
for Extract Entity. The initial data and type $D_1, T_1$ co-evolve into $D_2, T_2$ (adding
multiplicity) and a new value is then assigned to the array of assignees, resulting in $D_3$.
The difference from the earlier case is that the backward migration (from $D_3$ to $D'_3$)
generated for the type evolution (from $T_1$ to $T_2$) cannot be generated fully automatically.
In case of Extract Entity this was in principle possible.}

@@

If types, code and data can all be updated atomically together, then multiplicity change is
relatively straightforward  to handle. Existing scalar data can be trivially migrated to a
list, and the code using the data will need to change to accommodate the list data.
%
However, ongoing divergence makes multiplicity change much more difficult, and exposes
some general challenges. Different types may expose partial information, and writers using
those types have to operate without total knowledge of the information available in other types.
As a result, synchronizing writes across the types requires making difficult tradeoffs, such as
choosing to either preserve or destroy hidden data not visible to the writer.

@@

We considered live programming in the context of the Elm architecture in \S\ref{sec:elm},
where the challenge was to migrate run-time state (data) to a new type to keep the program running.
Work on language modeling offers a new perspective on the problem.
In the context of executable domain-specific modeling languages~\cite{klint2011easy}, one can see a
program in a domain-specific language as an instance of a metamodel (a static type) that
defines the AST structure of the domain-specific language.

@@

In the absence of live programming, the run-time \emph{types} and \emph{code} do not change while
running a program in a given domain-specific language and the run-time \emph{state} changes constantly.
Live programming, however, requires reconciling changes to the program with the induced run-time structures
of both the code and the state. This typically means that at a certain point (a quiescent point)
during execution, the code structures need to be updated (hot swapped), and the state needs to be
\textit{migrated}.

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-0.75em}
\includegraphics[width=14.5em]{figures/arr-model.png}
\vspace{-1.5em}
\end{wrapfigure}
\frameworkboxtitle{Code as Data Migrations}
We could see this challenge as type, code and data co-evolution as in the case of the Elm
architecture in \S\ref{sec:elm}, but a more interesting perspective is to see the
state machine definition as data $D_0$ conforming to a type $T$ defined by the metamodel.

@@

There has been extensive research on coordinating the evolution of interdependent layers of software, much of it guided by schema and types. But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system. Nevertheless it is important to understand these batch techniques to envision more interactive ones.

@@

\section{Conclusions}
We believe that effective future programming systems will need to tighten the feedback loop,
both in making programming more live and in making it more collaborative. One of the key obstacles
to making progress in this direction is the problem of \emph{type evolution}. When the structure
of a program evolves, code and data need to co-evolve to match the new structure.
%
The problem of type evolution has been explored in a wide range of sub-fields, but rarely in its
full complexity. Moreover, different sub-fields often use different language, making knowledge
transfer between them difficult.

>>>>

The first challenge problem that we present is live programming
of user interfaces based on the Elm architecture. In this model, a reactive web application is
structured in terms of current state and events that affect the state. The implementation then
consists of two functions called \texttt{update} and \texttt{render}:

@@

\noindent
The schema involved here are statically checked Elm types. The programming model works as follows:

@@

\noindent
Some systems that use the Elm architecture support hot reloading when the
implementation of the \texttt{render} or \texttt{update} function changes, but they typically
discard state and restart the application when the \texttt{State} type changes. A more effective
live programming system would allow live updates to the structure of the two types.

@@

\frameworkbox{
\begin{wrapfigure}{r}{12em}
\vspace{0em}
\includegraphics[width=12em]{figures/arr-steps.png}
\vspace{0em}
\end{wrapfigure}
\frameworkboxtitle{Primitive Schema Transformations}
In this case study, the evolving schema is the \texttt{State} type. When the type evolves,
the program logic (code) and the current live value of the type (data) needs to evolve
correspondingly. The challenge is easier if we see the evolution as a sequence of more primitive
transformations, such as those in Table~\ref{tbl:elmchanges}, that each has a corresponding
primitive code and data transformation. The primitive transformations may either be invoked
through a user interface, or inferred from textual code edits made by the programmer.
}

@@

\section{Entity Evolution in Data-Centric Systems}
If we see programming systems as stateful environments where programs co-exist with their persistent
data, we can gain valuable insights from research on database systems.
Although schema evolution is well-studied in databases (\S\ref{sec:related} Related Work), such past work
does not resolve all problems in our second challenge.

@@

\begin{figure}
\begin{adjustwidth}{-2em}{-3em}
\begin{subfigure}[b]{35em}\vspace{0pt}
  \sffamily
  \small
  \begin{tabular}{ |r|l|r|r|l|l|}
     \hline
     oid & item & quantity & ship\_date & customer\_name & customer\_address \\
     \hline \hline
     1 & Anvil & 1 & 2/3/23 & Wile E Coyote & 123 Desert Station \\
     \hline
     2 & Dynamite & 2 & & Daffy Duck & White Rock Lake \\
     \hline
     3 & Bird Seed & 1 & & Wile E Coyote & 123 Desert Station \\
     \hline
  \end{tabular}
  \caption{Spreadsheet recording orders with all associated information}
  \label{fig:db-orders}
\end{subfigure}
\hfill
\\[-0.5em]~\\
\begin{subfigure}[b]{20em}\vspace{0pt}
  \sffamily
  \small
  \begin{tabular}{ |r|l|l|}
    \hline
    cid & customer\_name & customer\_address \\
    \hline \hline
    1 & Wile E Coyote & 123 Desert Station \\
    \hline
    2 & Daffy Duck & White Rock Lake \\
    \hline
  \end{tabular}
  \caption{Customers table after schema evolution}
  \label{fig:db-customers}
\end{subfigure}
\hfill
\begin{subfigure}[b]{20em}\vspace{0pt}
  \sffamily
  \small
  \begin{tabular}{ |r|l|r|r|r|}
     \hline
     oid & item & quantity & ship\_date & cid \\
     \hline \hline
     1 & Anvil & 1 & 2/3/23 & 1 \\
     \hline
     2 & Dynamite & 2 & & 2  \\
     \hline
     3 & Bird Seed & 1 & & 1 \\
     \hline
    \end{tabular}
  \caption{Orders linking to Customers, after schema evolution}
 \label{fig:db-orders-link}
\end{subfigure}
\end{adjustwidth}
\vspace{0.25em}
\caption{Schema and data before and after performing Extract Entity}
\label{fig:db}
\end{figure}

@@

\frameworkbox{
\begin{wrapfigure}{r}{12em}
\vspace{-0.5em}
\includegraphics[width=12em]{figures/arr-db.png}
\vspace{-1em}
\end{wrapfigure}
\frameworkboxtitle{Equivalence in Schema Evolution}
%We described our conceptual framework in terms of types, code and data, but this case study shows that the same patterns appear elsewhere.
In database systems, the schema
$S_1$ changes into $S_2$ and the data $D_1$ need to transform to corresponding $D_2$.
Our framing shows that queries $Q_1$ also need to transform, which is not always addressed
in work on schema evolution. Extract/Absorb Entity operations are also interesting because
they transform an equivalent pair of schemas and so it should be possible to apply them
(and the corresponding data and query transformations) automatically.
}

@@

\section{Evolution of Computational Documents}
A different kind of schema evolution is present in programming systems based on documents.
The idea, dating back to Boxer \cite{diSessa86}, is to represent programs as documents
with embedded computations. In our model, document can contain simple formulas (akin to those
in Potluck~\cite{Litt2023} or spreadsheets) that may refer to other nodes of the document.
We assume that documents can be concurrently edited by multiple users in a local-first manner,
i.e., local edits are later synchronized with peers.

@@

   \vspace{0.5em}
   \textbf{Conference budget}\\
   Travel cost per speaker:\\
   \hspace*{2em} \$1200 \\
   Number of speakers: \\
   \hspace*{2em} \texttt{=COUNT(/table[id='speakers']/tbody/tr)} \\
   Travel expenses:\\
   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
   }}\caption{Geoffrey adopts changes from all three co-organizers.}
   \label{fig:conf-finfin}
\end{subfigure}
\end{adjustwidth}
\caption{
Initial version of the document (not shown) contains a list of three speakers (Goldberg, Hamilton,
Jennings). The first challenge involves merging a data edit
(Fig.~\ref{fig:conf-add}) with a schema and data edit (Fig.~\ref{fig:conf-tab}); the second involves
merging a schema and data edit (Fig.~\ref{fig:conf-tab}) with edit adding code (Fig.~\ref{fig:conf-budget}).
Formulas in the final version (Fig.~\ref{fig:conf-finfin}) are updated to match the new document structure.}
\label{fig:conf}
\end{figure}

@@

\begin{enumerate}
\item Jonathan adds an additional speaker to the list and sorts the list of speakers
  alphabetically by their first name. This changes data, but not the structure (schema) of
  the document (Figure~\ref{fig:conf-add}).
\item Tomas refactors the list into a table. He splits the single textual value
  into a name and an email (using a comma as the separator) and adds an additional column for
  tracking which of the organizers should contact the speaker (Figure~\ref{fig:conf-tab}).
  This changes the document structure (an implicit schema).
\end{enumerate}

@@

\frameworkbox{
\begin{wrapfigure}{r}{13em}
\vspace{-0.5em}
\includegraphics[width=13em]{figures/arr-datafork.png}
\vspace{-1em}
\end{wrapfigure}
\frameworkboxtitle{Merging Data and Code Edits}
In this example, the \emph{schema} is implicit and represents the structure of the document.
It involves two independent edits. In one (top right), the data $D_1$ is
changed to $D_2'$, but the schema $T_1$ remains unchanged. In the other (bottom left), the
schema $T_1$ and data $D_1$ co-evolve into $T_2$ and $D_2$ and the data is further changed
to $D_3$. The challenge is to merge the two edits and obtain jointly modified data $D_3'$
of schema $T_2$. In the next section, we will see the same pattern, but also involving
code edits. We assume the \emph{convergence} model in this section, but will discuss
the more challenging \emph{divergence} model later in the paper.}

@@

The schema and data evolution happens through a sequence of edits. Table~\ref{tbl:docchanges}
lists edits appearing in our examples (but not a complete list). A system may
use a structure editor that provides high-level commands for triggering those edits.
Recognising user intent from plain text editing of source code may make the challenge harder.

@@

\noindent
The challenge is, again, to merge the edits done independently by the co-organizers.
The challenge extends the previous one by adding code. The change to the document structure
(\emph{schema}) thus has to co-evolve with both the \emph{data} and the \emph{code}.
As shown in Figure~\ref{fig:conf-finfin}, the system needs to change the equation
\texttt{COUNT(/ul[id='speakers']/li)} to \texttt{COUNT(/table[id='speakers']/tbody/tr)}.
If the edit is recorded as a list of high-level edits shown in Figure~\ref{tbl:docchanges}
(wrapping element, changing their schema, etc.), then each edit needs to perform corresponding
edit to selectors in formulas in the document.

\begin{table}[t]
\begin{adjustwidth}{-2em}{-2em}
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Change & Tag of an element & Apply to all matching elements and affected formulas\\
Wrap & Selected element(s) & Wrap elements and update selectors in affected formulas  \\
Split & Selected element(s) & Split matching elements using a rule (regular expression)\\
Add & New item to a list & Add data without affecting document structure (schema)\\
Reorder & List items & Reorder data without affecting document structure (schema)\\
\bottomrule
\end{tabular}
\vspace{0.3em}
\caption{Changes to schema and data and how the programming system should respond}
\label{tbl:docchanges}
\end{adjustwidth}
\vspace{-1em}
\end{table}

\subsection*{Remarks: Liveness and Divergence}
The above challenge focuses on the local-first software scenario, but it can be extended to also
apply to live programming. If formulas in the document are evaluated on-the-fly, the system
needs to identify edits that invalidate some of the values (by changing the data that a formula
depends on, or by changing the equation). It should then automatically re-evaluate or invalidate
the computed results (requiring explicit re-evaluation). In this scenario, we distinguish between
\emph{permanent data} (information in the document) and transient \emph{data} (computed by code).
Whereas permanent data should be synchronized and co-evolve with the document structure (schema),
transient data can be discarded.

@@

\section{Divergence Control in Spreadsheets}
Schema evolution becomes more challenging when data, code, and Schema cannot all be updated
together in an atomic way. In these situations, different versions must coexist and interoperate
with one another. For example, edits done to data of a new schema must be applied to
data of an older schema, or code written against older schema must run against data of a newer schema.

@@

As a concrete example, we add the problem of divergence control to the
earlier Extract Entity challenge. Every month the orders department sends its spreadsheet to
the accounting department, which copies the data to its own version of the spreadsheet.
This version uses the same structure (schema), but includes additional code for financial tracking.
When the orders department migrates its spreadsheet to a database and extracts out customers,
the accounting department does not want to conform. They want to continue using their old
spreadsheet with their additional code.

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-1.5em}
\includegraphics[width=14em]{figures/arr-backport.png}
\vspace{-2.5em}
\end{wrapfigure}
\frameworkboxtitle{Bidirectional Data Migration}
The scenario adds the need for transferring data changes from a new schema to an older schema.
In the diagram on the right, we only show data and schema, ignoring the code added by the
accounting department. We start with data $D_1$ of schema~$T_1$. The accounting department
evolves the schema (via Extract Entity), obtaining data $D_2$ of schema $T_2$ using an automatically
generated forward transformation. It further modifies the data, obtaining $D_3$. When the
accounting department receives data $D_3$, it needs to migrate it back from schema $T_2$ to
$T_1$ using an automatically generated backward transformation. That is,
schema evolution needs to be accompanied with both forward and backward projection for data
(and code) migration.}

@@

\frameworkbox{
\begin{wrapfigure}{r}{14.5em}
\vspace{-0.5em}
\includegraphics[width=14.5em]{figures/arr-divergence.png}
\vspace{-1.5em}
\end{wrapfigure}
\frameworkboxtitle{Maintaining Divergent Variants}
The extra complexity in the above scenario arises from the fact that both schema and
code are forked and evolve independently. In one branch, original schema $T_1$ evolve
into $T_2$ and original code $C_1$ into $C_3$. In another branch, original schema $T_1$
evolve into $T'_2$ and original code $C_1$ into $C'_3$. The user then wants to merge
the independently done changes into a new version of schema and code, written as
$T_2 \diamond T'_2$ and $C_3 \diamond C'_3$. The diverging transformations may make
conflicting changes that would need to be resolved with assistance from the user.}

@@

The challenge of maintaining multiple divergent schema in programming systems can
be well illustrated using \emph{multiplicity change}, which is a common kind of schema change.
In this case a field changes from storing
a single value to storing multiple values. For example, we might start by storing a single
address for each contact in an address book, before realizing that we need to store multiple
addresses for a single contact. Alternatively, we might assign each todo in a list to one person before
realizing we want the ability to assign a todo to multiple people.

In a relational schema, solving this problem might require extracting a normalized table for
the linked entity; some of the challenges of this approach are covered in the Extract Entity
challenge. In this section we will instead consider a schema structure with support for arrays;
in this context, we can have a multiplicity change by turning a scalar value into an array value.
%
A change of multiplicity in the schema requires producing a transformation for both the corresponding
data and the corresponding code. The challenge becomes particularly difficult to handle when there
is ongoing divergence between multiple versions of the schema and data.

\challenge{Multiplicity Change in Schema}

@@

\frameworkbox{
\begin{wrapfigure}{r}{13em}
\vspace{-1em}
\includegraphics[width=13em]{figures/arr-backport.png}
\vspace{-2em}
\end{wrapfigure}
\frameworkboxtitle{Bidirectional Data Migration (Revisited)}
The diagram illustrating the scenario remains the same as in the case of divergence control
for Extract Entity. The initial data and schema $D_1, T_1$ co-evolve into $D_2, T_2$ (adding
multiplicity) and a new value is then assigned to the array of assignees, resulting in $D_3$.
The difference from the earlier case is that the backward migration (from $D_3$ to $D'_3$)
generated for the schema evolution (from $T_1$ to $T_2$) cannot be generated fully automatically.
In case of Extract Entity this was in principle possible.}

@@

If schema, code and data can all be updated atomically together, then multiplicity change is
relatively straightforward  to handle. Existing scalar data can be trivially migrated to a
list, and the code using the data will need to change to accommodate the list data.
%
However, ongoing divergence makes multiplicity change much more difficult, and exposes
some general challenges. Different schema may expose partial information, and writers using
those schema have to operate without total knowledge of the information available in other schema.
As a result, synchronizing writes across the schema requires making difficult tradeoffs, such as
choosing to either preserve or destroy hidden data not visible to the writer.

@@

We considered live programming in the context of the Elm architecture in \S\ref{sec:elm},
where the challenge was to migrate run-time state (data) to a new schema to keep the program running.
Work on language modeling offers a new perspective on the problem.
In the context of executable domain-specific modeling languages~\cite{klint2011easy}, one can see a
program in a domain-specific language as an instance of a metamodel (a static type AKA schema) that
defines the AST structure of the domain-specific language.

@@

In the absence of live programming, the run-time \emph{schema} and \emph{code} do not change while
running a program in a given domain-specific language and the run-time \emph{state} changes constantly.
Live programming, however, requires reconciling changes to the program with the induced run-time structures
of both the code and the state. This typically means that at a certain point (a quiescent point)
during execution, the code structures need to be updated (hot swapped), and the state needs to be
\textit{migrated}.

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-0.75em}
\includegraphics[width=14.5em]{figures/arr-model.png}
\vspace{-1.5em}
\end{wrapfigure}
\frameworkboxtitle{Code as Data Migrations}
We could see this challenge as schema, code and data co-evolution as in the case of the Elm
architecture in \S\ref{sec:elm}, but a more interesting perspective is to see the
state machine definition as data $D_0$ conforming to a type $T$ defined by the metamodel.

@@

There has been extensive research on coordinating the evolution of interdependent layers of software, much of it guided by schema and types. But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action and feedback in an interactive programming system. Nevertheless it is important to understand these batch techniques to envision more interactive ones.

@@

\section{Conclusions}
We believe that effective future programming systems will need to tighten the feedback loop,
both in making programming more live and in making it more collaborative. One of the key obstacles
to making progress in this direction is the problem of \emph{schema evolution}. When the structure
of a program evolves, code and data need to co-evolve to match the new structure.
%
The problem of schema evolution has been explored in a wide range of sub-fields, but rarely in its
full complexity. Moreover, different sub-fields often use different language, making knowledge
transfer between them difficult.
